Exploring the Role of Small Molecules in Biological Systems Using Network Approaches

### 2015-03-06 paper.tex
first draft done

Computational approaches in drug discovery, such as docking and
statistical QSAR, tend to focus on interactions between small
molecules and protein targets. Network representations and models play
a variety of roles in the \emph{in silico} drug discovery pipeline
ranging from visualization of heteregeneous datasets
\cite{Huan:2010cy,Kuhn:2010zr} to developing predictive models of drug
action \cite{Vina:2009oq,Folger:2011tx}. A necessary requirement for
network based analyses is the availability of network (or interaction)
data. This can come in many forms. Examples include protein-protein
interaction (PPI) databases such as BioGRID
\cite{Chatr-Aryamontri:2015yf} and STRING \cite{Franceschini:2013qa}
and drug-target databases such as DrugBank
\cite{Wishart:2008aa}. Bioactivity databases such as Pubchem and
ChEMBL \cite{Gaulton:2012jl} also provide useful information on
drug-target relationships, based on measured binding affinities.  In
addition to small molecules and proteins, databases such as PharmGKB
\cite{Thorn:2005aa} and MetaCore (a commercial product from Thompson
Reuters) integrate small molecules, protein and gene targets and the
published literature, providing a rich source of relationships between
these entities.

Using these databases, network methods can be applied to various
stages of the drug discovery pipeline. For example, target
identification can benefit from network analyses by characterizing the
role of a given target in the context of the larger interaction and
regulatory network. Thus targeting proteins that have a high degree of
connectedness may lead to significant side effects in other parts of
the system [REF] and can be the source of observed side effects. On
these lines, approaches to characterize toxicity have employed network
analyses \cite{Kim:2014xs,Zhang:2014hj}. While some approaches have
considered side effects [REF] others have considered the problem at
the genomic level by characterizing gene networks that are perturbed
by small molecules. Campaigns focused on combination therapies can
also benefit from a network representation and network inference
approaches \cite{Huang:2014by,Tang:2013lr} to identify target
combinations whose simultaneous inhibition leads to synergistic
effects. Mechanism of action studies can benefit from network analyses
by integrating multiple data types. An example of this was described
by Young et al \cite{Young:2008aa} that integrated high content
screening and ligand-target prediction models to construct network
representations of MoA hypotheses. Network approaches have also been
proposed for SAR analysis \cite{Iyer:2011ij,Wawer:2008aa}, usually at
the lead optimization stage of the discovery pipeline. Drug
repositioning efforts \cite{Dudley:2011dn,Gottlieb:2011rx} benefit
from network analyses such as community detection algorithms
\cite{girvan2002community}.

@@

% While the \rcdk package and its dependencies provide general small
% molecule handling functionality, a number of other packages provide
% more specific functionality, related to small molecules. We have
% previously mentioned DrugComboRanker \cite{huang2014drugcomboranker}
% and MANTRA \cite{iorio2010discovery}.

@@

In this chapter we have attempted to provide an overview of how
networks have been employed to analyze and visualize chemical
data. The fact that chemical structures themselves are graphs allows
one to apply a number of techniques from graph theory to characterize
chemical structure (such as the many graph invariants). However, with
the advent of large collections of chemical structures (from databases
such as Pubchem or high throughput screening programs) along with
associated data such as bioactivity and target annotations, a number
of workers have attempted to analyze the properties of such
collections using network techniques.

It is interesting to note that many applications of network methods to
chemical collections focus on visualization of the chemical data,
rather than using the network structure to draw insight into the
underlying chemical of biological phenomena. This is not to say that
network visualizations of chemical data are lacking. In fact, such
visualizations allow one to encode multiple properties of objects
(i.e., nodes) and relationships (i.e., edges) in a single visual form.

@@

>>>>

Computational approaches in drug discovery, such as docking and
statistical QSAR, tend to focus on interactions between small
molecules and protein targets. By definition they do not consider the
surrounding functional environment of interacting proteins and related
pathways. Network representations on the other hand are capable of
capturing the functional context of individual molecular entities and
network models play a variety of roles in the \emph{in silico} drug  
discovery pipeline ranging from visualization of heteregeneous
datasets \cite{Huan:2010cy,Kuhn:2010zr} to developing predictive
models of drug action \cite{Vina:2009oq,Folger:2011tx}. A necessary
requirement for network based analyses is the availability of network
(or interaction) data. This can come in many forms. Examples include
protein-protein interaction (PPI) databases such as BioGRID
\cite{Chatr-Aryamontri:2015yf} and STRING \cite{Franceschini:2013qa}
and drug-target databases such as DrugBank
\cite{Wishart:2008aa}. Bioactivity databases such as Pubchem and
ChEMBL \cite{Gaulton:2012jl} also provide useful information on
drug-target relationships, based on measured binding affinities.  In
addition to small molecules and proteins, databases such as PharmGKB
\cite{Thorn:2005aa} and MetaCore (a commercial product from Thompson
Reuters) integrate small molecules, protein and gene targets and the
published literature, providing a rich source of relationships between
these entities.

Using these databases, network methods can be applied to various
stages of the drug discovery pipeline. For example, target
identification can benefit from network analyses by characterizing the
role of a given target in the context of the larger interaction and
regulatory network. Thus targeting proteins that have a high degree of
connectedness (as measured by essentiality and centrality) may lead to
significant side effects in other parts of the system and can be the
source of observed side effects \cite{Wang:2013gn}. On these lines,
approaches to characterize toxicity have employed network analyses
\cite{Kim:2014xs,Zhang:2014hj}. While some approaches have considered
side effects [REF] others have considered the problem at the genomic
level by characterizing gene networks that are perturbed by small
molecules. Campaigns focused on combination therapies can also benefit
from a network representation and network inference approaches
\cite{Huang:2014by,Tang:2013lr} to identify target combinations whose
simultaneous inhibition leads to synergistic effects. Mechanism of
action studies can benefit from network analyses by integrating
multiple data types. An example of this was described by Young et al
\cite{Young:2008aa} that integrated high content screening and
ligand-target prediction models to construct network representations
of MoA hypotheses. Network approaches have also been proposed for SAR
analysis \cite{Iyer:2011ij,Wawer:2008aa}, usually at the lead
optimization stage of the discovery pipeline. Drug repositioning
efforts \cite{Dudley:2011dn,Gottlieb:2011rx} benefit from network
analyses such as community detection algorithms
\cite{girvan2002community}.

While drugs can range from small molecules to antibodies and from pure
compounds to mixtures, the following sections focus on how network
approaches have been used to characterize and model the interactions
of small molecules with biological systems. We group such interactions
based on the nature of the interacting entities and provide a brief
overview of recent work on how network representations have been
employed to analyze them.

@@

@@

In this chapter we have summarized how
networks have been employed to analyze and visualize small molecule
data (structures, interactions, activity) in the context of protein
and gene targets. The fact that chemical structures themselves are graphs allows
one to apply a number of techniques from graph theory to characterize
chemical structure (such as the many graph invariants). However, with
the advent of large collections of chemical structures (from databases
such as Pubchem or high throughput screening programs) along with
associated data such as bioactivity and target annotations, a number
of workers have attempted to analyze the properties of such
collections using network techniques.

It is interesting to note that many applications of network methods to
chemical collections focus on visualization of the chemical data,
rather than using the network structure to draw insight into the
underlying chemical or biological phenomena. This is not to say that
network visualizations of chemical data are lacking. In fact, such
visualizations allow one to encode multiple properties of objects
(i.e., nodes) and relationships (i.e., edges) in a single visual
form. However, in many cases, studies tend to report structural
characterizations (average vertex degree, centrality values and so on)
but do not make a persuasive link between these and the underlying
chemistry and biology.

@@

In conclusion, network representations of molecular entities and their
interactions are a powerful tool that captures (possibly) large and
heterogenous datasets. Network approaches have been employed at
various stages of the drug discovery pipeline ranging from target
identification to SAR analysis. While one could argue that many
results from network approaches could be achieved by simpler tabular
queries, it is evident that in a number of cases, the integrative
nature of network methods provides a novel view of the data and opens
up alternative avenues of investigation.

\section{Acknowledgements}
\label{sec:acknowledgements}

We would like to thank Dac-Trung Nguyen and Tyler Peryea for fruitful
discussions related scaffold and scaffold-document networks.
